---
title: "Vector API Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector API Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

Beginning at version 2.0, **gdalraster** includes bindings to a substantial subset of the GDAL Vector API. The vector bindings provide an R implementation of the [GDAL Vector Data Model](https://gdal.org/en/stable/user/vector_data_model.html). Readers are encouraged to first review that document for background on the underlying object model and how it relates to usage of the API in **gdalraster**.

The prefix [OGR](https://gdal.org/en/stable/faq.html#what-is-this-ogr-stuff) is used in GDAL with class names, command-line utilities, etc., that refer to vector-specific components of the library. In **gdalraster**, OGR will be used throughout the documentation and as a prefix on function names, consistent with the style of the source library. There is also a certain amount of unification of the raster and vector APIs in GDAL. In **gdalraster**, we will also refer to the "GDAL Vector API" and use `GDALVector` as a class name. To support the vector API, **gdalraster** 2.0 builds on its already existing bindings to portions of GDAL's Spatial Reference System and Geometry APIs.

## Rationale

As an implementation of GDAL's Raster and Vector Data Models, **gdalraster** is not intended as a direct alternative to existing R packages such as **sf** and **terra** that primarily implement higher-level analytical data models. **gdalraster** is primarily focused on API-level bindings, attempting to expose most of the underlying C++ library to R. More direct access to GDAL's I/O capabilities may benefit workflows that deal with source data provided on a continuum ranging from local storage and in-memory formats to database servers and cloud systems, often involving various compression formats, SQL dialects, etc. Modularized, low-level access to GDAL/OGR functionality should be of interest to R developers creating higher-level interfaces, where the analytical data models may not have been designed with their use cases in mind. API bindings facilitate code translation and reuse which can benefit hybrid environments where GDAL is used from multiple languages. Long-term stability of bindings at the library level is highly desirable in production scenarios, and is a major goal for the package. It is intended that **gdalraster** should complement and interoperate with R's diverse and evolving spatial ecosystem.

## Vector API components

This section describes the organization of the vector components in **gdalraster**, linking to package documentation for extended text on specific topics.

### Core bindings

Core functionality for vector is provided by an exposed C++ class `GDALVector`, along with a set of stand-along functions with `ogr_()` prefix.

#### Class `GDALVector`

`GDALVector` is a C++ class exposed in R via [**Rcpp** modules](https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-modules.pdf). It encapsulates a single OGRLayer and the GDALDataset that contains it. An object of class `GDALVector` persists an open connection to the dataset, and exposes methods to: retrieve layer information, set attribute and spatial filters, traverse and read feature data by traditional row-based cursor (including an analog of `DBI::dbFetch()`), read via column-oriented Arrow Array stream, write new features in a layer, edit/overwrite existing features, upsert, and delete features, and perform data manipulation within transactions. The `?GDALVector` documentation gives a full list of object methods under Usage, and descriptions of each under Details.

#### OGR stand-alone functions

Several of the stand-alone `ogr_()` functions are grouped under the documentation topics `?ogr_manage` and `?ogr_define`. The `ogr_manage` functions can be used to:

* create new vector datasets from scratch
* test for existence of datasets, layers and fields
* test dataset and layer capabilities
* create new layers in an existing dataset
* delete layers
* create new attribute and geometry fields on an existing layer
* rename and delete fields
* edit data with SQL statements

`ogr_define` provides documentation and helper functions for defining feature classes. An OGR feature class definition (a.k.a. layer definition) is modeled in R as a named list containing zero or more attribute field definitions, along with one or more geometry field definitions. Specifications of the the list structures for these definitions are given in `ogr_define`. The associated helper functions make it easy to create new layer definitions from scratch or modify an existing definition. A layer definition is convenient but not required for creating a new vector dataset, and/or a new layer within an existing dataset, using `ogr_ds_create()` / `ogr_layer_create()`.

OGR facilities for vector geoprocessing are available in `ogr_proc()`. This function performs the following GIS overlay operations: Intersection, Union, SymDifference, Identity, Update, Clip and Erase (https://en.wikipedia.org/wiki/Vector_overlay). `ogr_proc()` is basically an R port of the command-line utility[`ogr_layer_algebra`](https://gdal.org/en/stable/programs/ogr_layer_algebra.html#ogr-layer-algebra) which is included in the GDAL Python bindings. Both of these are interfaces to library functions via OGR API calls.

`ogrinfo()` and `ogr2ogr()`, provide R wrappers of the GDAL command-line utilities [`ogrinfo`](https://gdal.org/en/stable/programs/ogrinfo.html) and [`ogr2ogr`](https://gdal.org/en/stable/programs/ogr2ogr.html), respectively. These functions support all of the command-line arguments described in the GDAL documentation, providing a powerful set of capabilities for obtaining information about an OGR-supported data source, converting vector data between file formats, and potentially editing data with SQL statements. `ogr_reproject()` is a convenience wrapper around `ogr2ogr()` for reprojection of vector layers with user friendly interface and comprehensive documentation.

### Supporting APIs

#### Spatial Reference Systems

Bindings to a subset of GDAL's Spatial Reference System API are provided by a set of stand-alone `srs_()` functions. These support conversion of spatial reference system definitions in various formats to OGC Well Known Text (WKT) representation (e.g., EPSG codes, well known names such as NAD27, NAD83, WGS84, etc., PROJ.4 definitions, PROJJSON, and others, as well as between different versions of OGC WKT, see `?srs_convert`). Several functions under the `?srs_query` topic provide various information about a given SRS definition and allow comparing definitions for equality.

#### Geometry

Bindings to a significant subset of GDAL's Geometry API are implemented in a set of functions prefixed `g_()`. Many of these functions are built on the [GEOS library](https://libgeos.org/) (i.e., using GEOS via GDAL headers). The geometry functions in **gdalraster** are flexible in that the input geometries, and corresponding output, may be any of a single [WKT string](https://libgeos.org/specifications/wkt/), a character vector of WKT strings, a single `raw` vector of [WKB](https://libgeos.org/specifications/wkb/), or a list of WKB `raw` vectors. The convenience function `g_wk2wk()` will convert its input from one to the other accordingly. Factory functions (`?g_factory`) support creating basic geometry types from point data (vertices) and building container geometry types from sub-geometries. Capabilities of the geometry bindings include inspection (`?g_query`), binary predicates (`?g_binary_pred`), binary operations (`?g_binary_op`), unary operations (e.g., `g_buffer()`, `g_simplify()`), measurements (`?g_measures`), and several utilities (e.g., `g_coords()`, `g_make_valid()`, `g_swap_xy()`, `g_transform()`). Convenience functions for working with bounding boxes are also provided (e.g., to/from WKT, intersect/union, and `bbox_transform()` to a different projection).

#### Virtual Systems Interface

Bindings to the GDAL's Virtual Systems Interface (VSI) implement standard file system operations abstracted for URLs, cloud storage services, Zip/GZip/7z/RAR, in-memory files, as well as "regular" local file systems. This provides a single interface for operating on file system objects, that works the same for any storage backend. The `vsi_()` functions have general utility not limited to operating on spatial data sources.

#### General data management

Existing data management functions that operate on both raster and vector data sources include `addFilesInZip()` (create/append to a potentially [Seek-Optimized ZIP](https://github.com/sozip/sozip-spec) file), `deleteDataset()`, `identifyDriver()` and `inspectDataset()`.

## Workflow considerations

Creation of new vector datasets, and schema modification in existing datasets, are performed with the `ogr_()` stand-alone functions. These are "one-off" operations that attempt to open the dataset with update access, perform modifications, and then close the dataset. Objects of class `GDALVector` are used for reading and writing feature data. A `GDALVector` object provides a persistent connection to a layer and the vector dataset that contains it. That is, once instantiated, a `GDALVector` object represents a live connection to the data source until its `close()` method is called. The connection may be read-only (by default), or may be with update access allowing insert of new features, modifying exiting features, and deleting features. Currently, class `GDALVector` does not provide methods for modifying the layer schema.



High-level workflow for using **gdalraster** with vector data: distinguishing among phases that involve creation and management of vector data sources, schema management, layer access and data retrieval, writing/modifying data at the individual feature level, and vector geoprocessing on whole layers.



### Data retrieval


https://en.wikipedia.org/wiki/Cursor_(databases)


OGR field types are returned as the following R types (`NA` for OGR NULL values). When retrieving a batch of features as a data frame (`"OGRFeatureSet"`), some field types will be contained in a data frame list column as indicated:

* **OFTInteger**: `integer` value
* **OFTInteger subtype OFSTBoolean**: `logical` value
* **OFTIntegerList**: vector of `integer` (list column)
* **OFTInteger64**: `numeric` value carrying the `"integer64"` class attribute
* **OFTInteger64 subtype OFSTBoolean**: `logical` value
* **OFTInteger64List**: vector of `bit64::integer64` (list column)
* **OFTReal**: `numeric` value
* **OFTRealList**: vector of `numeric` (list column)
* **OFTString**: `character` string
* **OFTStringList**: vector of `character` strings (list column)
* **OFTDate**: `numeric` value of class `"Date"`
* **OFTDateTime**: `numeric` value of class `"POSIXct"` (millisecond accuracy)
* **OFTTime**: `character` string (`"HH:MM:SS"`)
* **OFTBinary**: `raw` vector (list column, `NULL` entries for OGR NULL values)

By default, geometries are returned as WKB `raw` vectors in a data frame list column. The per-object setting `$returnGeomAs` can also be set to one of `"WKB_ISO"`, `"WKT"`, `"WKT_ISO"`, or `"NONE"`. Omitting the geometries (e.g., by setting `lyr$returnGeomAs <- "NONE"`) may be beneficial for performance and memory usage when access only to feature attributes is required.





## Code examples

Code examples for common tasks including: create new vector data sources, define and create new layers in a data source, connect to layers, filter and read feature data, write new data, etc.

Note that parentheses around statements are used throughout the vignette only as a shortcut to display the assigned value.

### Inspect data sources

Data sources can be files, relational database management systems (RDBMS), directories of many files, or even remote web services depending on the format driver being used. However, the data source name (DSN) is always a single string which might be the file path, database connection string, URL, etc.

The following examples use a GeoPackage file included in **gdalraster**. The `system.file()` command is only used to obtain the path to the package sample files. The file ynp_features.gpkg is compressed using SOZip and will be read directly from the .zip archive without decompressing first. A prefix is added to the file path ([/vsizip/](https://gdal.org/en/stable/user/virtual_file_systems.html#vsizip-zip-archives) in this case) which specifies a GDAL Virtual File System handler. A file system handler provides access to less standard file types such as in-memory, compressed (.zip, .gz, .tar, .tar.gz), encrypted, standard input and output (STDIN, STDOUT), files stored on network (publicly accessible, or in private buckets of commercial cloud services), etc.

```{r}
library(gdalraster)

# get path to the Yellowstone National Park sample dataset
f <- system.file("extdata/ynp_features.zip", package = "gdalraster")

# add the VSI prefix 
(zf <- file.path("/vsizip", f))

# list files in the zip archive
vsi_read_dir(zf)

# VSI path to the GPKG file
(zf_gpkg <- file.path(zf, "ynp_features.gpkg"))
```

Note that it is possible to chain multiple file system handlers. The [/vsicurl/](https://gdal.org/en/stable/user/virtual_file_systems.html#vsicurl-http-https-ftp-files-random-access) prefix specifies a file system handler that allows on-the-fly random reading of files available through web protocols, without prior download of the entire file. For example, a zip file residing on a web server instead of the local file system could be accessed with a path like:

```
/vsizip//vsicurl/https://www.example.com/path/to/file.zip
```

Support for SOZip is available with GDAL >= 3.7. The function `vsi_get_file_metadata()` can be used to validate an SOZip file and obtain compression information. Otherwise, SOZip is fully backward compatible and works as a regular .zip file.

```{r}
if (gdal_version_num() >= gdal_compute_version(3, 7, 0)) {
  print("SOZip metadata for ynp_features.gpkg:")
  print(vsi_get_file_metadata(zf_gpkg, domain="ZIP"))
} else {
  print("SOZip support requires GDAL >= 3.7")
}
```

`inspectDataset()` returns information about the format and content of a data source that may contain raster and/or vector data.

```{r}
inspectDataset(zf_gpkg)
```

OGR functions can be also be used to inspect a vector data source.

```{r}
# test for existence of a vector data source with at least read access
ogr_ds_exists(zf_gpkg)
# note that update of an existing zipped GPKG file is not supported
ogr_ds_exists(zf_gpkg, with_update = TRUE)

# list the vector layers
ogr_ds_layer_names(zf_gpkg)

# ogrinfo() requires GDAL >= 3.7
if (gdal_version_num() >= gdal_compute_version(3, 7, 0)) {
  cat("List of layers found in the data source:\n")
  cat(ogrinfo(zf_gpkg, cout = FALSE))
  cat("\n")
  cat("Detailed information about a specific layer:\n")
  cat(ogrinfo(zf_gpkg, "points_of_interest", cout = FALSE))
}
```

### Manage data sources and schemas

Here we copy the .gpkg file from the zip archive to an in-memory file that is writable, and make several modifications. The [/vsimem/](https://gdal.org/en/stable/user/virtual_file_systems.html#vsimem-in-memory-files) file handler allows a block of memory to be treated as files. It is useful for temporary storage and is generally fast to access. The code below also demonstrates testing capabilities at the dataset or layer level before attempting to perform modifications. This will not be needed in all cases but might be used in code that is required to handle input datasets in a general way.

```{r}
# copy ynp_features.gpkg from the zip file to an in-memory file
mem_gpkg <- "/vsimem/tmp/ynp_features.gpkg"
ogr2ogr(zf_gpkg, mem_gpkg)

vsi_read_dir("/vsimem/tmp")

# confirm it's writable
ogr_ds_exists(mem_gpkg, with_update = TRUE)

# rename a layer
if (ogr_layer_test_cap(mem_gpkg, "points_of_interest")$Rename) {
  ogr_layer_rename(mem_gpkg, "points_of_interest", "ynp_poi")
} else {
  stop("layer does not have Rename capability")
}

# delete a layer
if (ogr_ds_test_cap(mem_gpkg)$DeleteLayer) {
  ogr_layer_delete(mem_gpkg, "roads")
} else {
  stop("dataset does not have DeleteLayer capability")
}

ogr_ds_layer_names(mem_gpkg)

# manage fields on a layer
ogr_layer_field_names(mem_gpkg, "ynp_poi")

# delete a field
if (ogr_layer_test_cap(mem_gpkg, "ynp_poi")$DeleteField) {
  ogr_field_delete(mem_gpkg, "ynp_poi", "objectid")
} else {
  stop("layer does not have DeleteField capability")
}

# rename fields
if (ogr_layer_test_cap(mem_gpkg, "ynp_poi")$AlterFieldDefn) {
  ogr_field_rename(mem_gpkg, "ynp_poi", "poiname", "poi_name")
  ogr_field_rename(mem_gpkg, "ynp_poi", "poitype", "poi_type")
} else {
  stop("layer does not have AlterFieldDefn capability")
}
```

As a simple example for illustration, we populate a new field that flags geothermal features.

```{r}
# create a new field
if (ogr_layer_test_cap(mem_gpkg, "ynp_poi")$CreateField) {
  ogr_field_create(mem_gpkg, "ynp_poi", "is_geothermal",
                   fld_type = "OFTInteger",
                   fld_subtype = "OFSTBoolean")
} else {
  stop("layer does not have CreateField capability")
}

# edit data with SQL
sql <- "UPDATE ynp_poi SET is_geothermal = 
          CASE
            WHEN poi_type IN ('Basin', 'Geyser') THEN TRUE
            ELSE FALSE
          END"
ogr_execute_sql(mem_gpkg, sql)

ogr_layer_field_names(mem_gpkg, "ynp_poi")
```

To read back from the modified data source we need to instantiate objects of class `GDALVector`. More detail on working with `GDALVector` objects is given in the examples below for data retrieval. For now, note that the function `ogr_execute_sql()` returns an object of class `GDALVector` for SQL statements that return a result set.

```{r fig.width=6, fig.height=4, dev="png"}
# read and display the geothermal features
sql <- "SELECT poi_name, geom FROM ynp_poi WHERE is_geothermal = TRUE"
lyr <- ogr_execute_sql(mem_gpkg, sql)
lyr$getFeatureCount()

# retrieve all features in the result set (cf. DBI::dbFetch())
feat_set <- lyr$fetch(-1)
head(feat_set)

# plot the park boundary, this layer contains a single polygon feature
GDALVector$new(mem_gpkg, "ynp_bnd")$getNextFeature() |>
  plot(col = "wheat", xlab = "longitude", ylab = "latitude",
       main = "YNP Geothermal Features")

# add point locations of the geothermal features
plot(feat_set, pch = 19, col = "steelblue1", add = TRUE)
```

For clean up, the `$close()` method should be called on the layer object opened with `ogr_execute_sql()`. We also delete the temporary in-memory file.

```{r}
lyr$close()

# delete a data source
vsi_unlink(mem_gpkg)  # delete a single file
# or, deleteDataset(mem_gpkg)
```

### Traditional row-based data retrieval

These examples will use the original sample dataset for YNP features. The DSN is formed with the /viszip/ prefix and full path to ynp_features.gpkg contained in the zip archive. An object of class `GDALVector` is generated with a call to `new()`. From the examples above, we know that the GeoPackage contains three vector layers, so we also pass a layer name argument to the object constructor. Note that if the layer argument is omitted, the constructor will attempt to open the first layer found by index. This abbreviated form of the constructor would generally be used for convenience only with single-layer file formats (e.g., ESRI Shapefiles, FlatGeoBuf), or with a GeoPackage file known to contain only a single layer.

Here we instantiate a `GDALVector` object for the park boundary layer and retrieve information about it.

```{r}
f <- system.file("extdata/ynp_features.zip", package = "gdalraster")
ynp_dsn <- file.path("/vsizip", f, "ynp_features.gpkg")

# the park boundary layer containing a single feature
(bnd <- new(GDALVector, ynp_dsn, "ynp_bnd"))

bnd$getFeatureCount()

# spatial reference definition as WKT string
bnd$getSpatialRef()

# xmin, ymin, xmax, ymax
bnd$bbox()

# structure of the layer definition (a.k.a. feature class definition)
bnd$getLayerDefn() |> str()
```

Traditional row-level access to feature data with OGR is cursor-based, reading one feature (row) at a time. When starting with a new layer object, the cursor is positioned at the beginning and we can iterate through features with calls to `$getNextFeature()`. This method returns an individual feature as a named list of fields and their values, and moves the cursor forward one row. It will return `NULL` when we run out of features. The park boundary layer contains a single polygon feature. The list object for an individual feature carries the `"OGRFeature"` class attribute with S3 methods for `print()` and `plot()`.

```{r}
bnd_feat <- bnd$getNextFeature()
str(bnd_feat)

# no more features
bnd$getNextFeature()

bnd$close()
```

The cursor can be reset to the beginning of the layer at any time with a call to `$resetReading()`.

Class `GDALVector` also provides a `$fetch()` method to retrieve the next `n` features from the layer and return them as a data frame. `$fetch()` is an analog of `DBI::dbFetch()` with basically identical calling semantics, where the vector layer itself is the equivalent of a **DBI** result set. `$fetch()` can be used to get the entire feature set for a layer, one batch of features at a time, or the remaining features from the current cursor position. Fetching zero features is also possible to retrieve the structure of the layer as a 0-row data frame with fully typed columns. As with `$getNextFeature()`, only forward paging is supported.

Passing `n = -1` or `n = Inf` will retrieve all features from the beginning (with an automatic call to `$resetReading()`). Passing `n = NA` is supported and will retrieve the remaining features from the current cursor position. Otherwise, features can be fetched progressively by passing a whole number (`integer` or `numeric`) as the `n` argument. If more features than available are fetched, the result is returned in full without warning. If fewer features than requested are returned, further fetches will return a data frame with zero rows.

Note that it is possible to set an attribute and/or spatial filter prior to retrieving data to restrict the feature set returned. It is also possible to set specific fields as ignored (or to specify the list of desired fields). This includes potentially omitting the geometry column when access only to feature attributes is needed. Limiting the number of fields returned may save some processing time and/or bandwidth.

Here we read the roads layer after setting an attribute filter to select only the publicly accessible roads. We call `$fetch()` with `n = -1` to read from the beginning and retrieve the full feature set. The data frame returned by `$fetch()` carries the `"OGRFeatureSet"` class attribute with S3 methods for `print()` and `plot()`.

```{r fig.width=6, fig.height=4, dev="png"}
# an object for the roads layer
(roads <- new(GDALVector, ynp_dsn, "roads"))

roads$getFeatureCount()

roads$getFieldNames()

roads$setAttributeFilter("opentopubl = 'Yes'")

roads_featset <- roads$fetch(-1)
nrow(roads_featset)

head(roads_featset)

plot(bnd_feat, col = "wheat", xlab = "longitude", ylab = "latitude",
     main = "YNP Public Roads")

plot(roads_featset, col = "slategray", lwd = 2, add = TRUE)

roads$close()
```

Fetching in batches may be advantageous when dealing with large datasets.

```{r}
poi <- new(GDALVector, ynp_dsn, "points_of_interest")

poi$getFeatureCount()

# access features progressively in batches
batch <- 0
while (TRUE) {
    poi_featset <- poi$fetch(500)
    if (nrow(poi_featset) == 0) break
    cat("batch", batch <- batch + 1, ":", nrow(poi_featset), "features\n")
    # process batch
    # ...
}

poi$close()
```

### Reading with OGR Arrow ArrayStream interface

It is also possible to retrieve features in a column-oriented memory layout using OGR's Arrow ArrayStream interface (GDAL >= 3.6). Performance advantages may be substantial with vector format drivers that provide specialized implementations (e.g., Arrow, Parquet, FlatGeoBuf, GeoPackage). Drivers that have a specialized implementation advertise the `FastGetArrowStream` layer capability.

The `$getArrowStream()` method of class `GDALVector` exposes an ArrayStream on the layer as a `nanoarrow_array_stream` object. The **nanoarrow** package provides functionality to consume the ArrayStream and import to R data structures.

The default batch size for the ArrayStream is 65,536 features, but this can be changed on a per-object basis by assigning a value to the `GDALVector` writeable field `$arrowStreamOptions` (a character vector of `"NAME=VALUE"` pairs).

```{r}
# re-open the roads layer with the required argument for type of access
roads$open(read_only = TRUE)
roads$resetReading()

# Expose an Arrow array stream (requires GDAL >= 3.6)
if (gdal_version_num() >= gdal_compute_version(3, 6, 0)) {

  cat("FastGetArrowStream layer capability:\n")
  roads$testCapability()$FastGetArrowStream |> print()
  cat("\n")

  # change the default ArrowStream options
  roads$arrowStreamOptions <- c("INCLUDE_FID=NO")

  stream <- roads$getArrowStream()
  cat("Object returned by getArrowStream():\n")
  print(stream)
  cat("\n")

  # get the array schema if needed
  schema <- stream$get_schema()

  # read a batch of features (default batch size is 65,536)
  batch <- stream$get_next()

  # disable a warning that can be safely ignored here
  options(nanoarrow.warn_unregistered_extension = FALSE)

  cat("Array structure to data frame provided by nanoarrow:\n")
  batch_df <- as.data.frame(batch)
  head(batch_df) |> print()
  cat("\n")

  cat("The last batch is NULL: stream$get_next()\n")
  stream$get_next() |> print()
  cat("\n")

  # release the stream when finished
  stream$release()
  cat("Stream released:\n")
  print(stream)
  cat("\n")

  # the geom column is a list column of WKB raw vectors
  cat("Road length calculated with Geometry API g_*() functions:\n")
  geom_utm12 <- g_transform(batch_df$geom,
                            srs_from = roads$getSpatialRef(),
                            srs_to = "EPSG:26912")

  batch_df$rdlength <- g_length(geom_utm12)
  head(batch_df) |> print()
}

roads$close()
```
